Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-08-15 manuscript/bmc_article.tex
Spearman correlation estimates array<->RNAseq

The number of sequence read mappings obtained for individual
replicates ranged from 11,479,260 (sample NMRI\_sporozoites\_rep2) to
145,033,326 (NMRI\_2ndInf\_5dpi\_rep1).

As samples and idividual replicates were sequenced in batches to
different depth and using different instrumentation (Table 1) we
performed quality controls confirming the abscence of batch effects
influencing analysis and quality of our results (additional file xyc).

The results of these multivariate analyses in relation to biological
interpretation are described further below.

@@

A maximum of 86\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of
0.0326\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be
assigened to the \textit{E. falciformis} genome in samples considered
infected.

@@

On the other end of the spectrum mice not previously immunized
featured an mRNA ``holo-transcriptome'' at 7dpi with the majority of
mRNA originating from the parasite as opposed to the host (Figure
1). The overall mRNA output of the sampled caecum tissue at this point
of infection is dominated by the parasite with proportional mRNA
abundance between XXX and XXX %

The total number of sequencing reads obtained as well as those reads
mapped to the transcriptome of either \textit{E. falciformis} and the
mouse are indicated in Table 1 for all replicates.

@@

The observed changes between controls and 7dpi were in agreement with
previously published microarry data. Fold-Change data obtained from
mice at 6dpi on Agilent microarrays (Schmidt et al 2012) or 7dpi (our
RNAseq data) agains uninfected controls shows a strong correlateion
(Spearman's \rho = 333; $R^2$ from a linear model = 444 ; figure 2
a). Considering both and technical differences between the tow methods
and the biological changes along the lifecycle this confirms the
adequacy of our methods for assessing the host transcriptome.

XX mouse genes were considered differently expressed (DE; FDR<1\%)
between uinfected and 3dpi, XX genes between controls and 5dpi, xxx
genes between 0dpi and 7dpi. This lead to combined set of xxxx genes
reacting to infection (see figure 2 b).

We detected no distinct response early in infection.  When testing for
differences between 3dpi and later infection (5dpi and 7dpi) DE genes
were simply a subset of genes DE between respective infections and
controls.

To further validate these patterns we performed hierarchical
clustering on mouse mRNAs DE at different dpi (union of DE
genes).

%% Samples from day 7 p.i. form a separate cluster (NMRI mice only), not
%% distinguishing between first and challenge infection. For these
%% samples, the pattern is most pronounced in gene cluster 1 (up), and 4
%% and 6 (down). Within cluster 1, the second replicate from the
%% challenge infection has a deviant profile (compare mouse data in
%% Figure 2). Distinct groups of genes also define sporozoites (cluster 5
%% up) and oocysts (cluster 3 up, cluster 6 down). mRNA profiles on days
%% 3 and 5 p.i. from all three mouse strains cluster together. These
%% samples are distinct from oocysts, NMRI day 7 p.i., and
%% sporozoites. Sporozoites cluster closer with days 3 and 5 than with
%% oocysts or day 7.

%%% Test _this_:
The more ponounced changes later in infection are is likely reflecting
a lag time in detection before the parasite reaches sufficient
intensity during asexual expansion (schizogony). Those changes are
mainly related to the onset of an adaptive immune response.

>>>>

The number of total read mappings obtained for individual replicates
ranged from 11,479,260 (sample NMRI\_sporozoites\_rep2) to 145,033,326
(NMRI\_2ndInf\_5dpi\_rep1). As samples and idividual replicates were
sequenced in batches to different depth and using different
instrumentation (Table 1) we performed quality controls confirming the
abscence of batch effects influencing analysis and quality of our
results (additional file xyc). The results of these multivariate
analyses in relation to biological interpretation are described
further below.

@@

A maximum of 86\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of
0.0326\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be
assigened to the \textit{E. falciformis} genome in samples considered
infected (table 1).

@@

On the other end of the spectrum mice not previously immunized
featured an mRNA ``holo-transcriptome'' at 7dpi with the majority of
mRNA originating from the parasite as opposed to the host (Figure
1). The overall mRNA output of the sampled caecum tissue at this point
of infection is dominated by the parasite with proportional mRNA
abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and even 92\% (in
NMRI\_1stInf\_7dpi\_rep1).

The total number of sequencing reads obtained as well as those reads
mapped to the transcriptome of either \textit{E. falciformis} and the
mouse are indicated in Table 1 for all replicates. We additonally
assessed the number of genes considered ``expressed'', further
analysis and discussion considers the possibility that especially for
\textit{E. falciformis} the number of reads might have been
insufficient to fully normalize the dataset of the samples with the
lowest representation against those with the hightest.

@@

The observed changes between controls and 7dpi were in agreement with
previously published microarry data. Fold-Change data obtained from
mice at 6dpi on Agilent microarrays (Schmidt et al 2012) or 7dpi (our
RNAseq data) against uninfected controls shows a strong correlateion
(Spearman's \rho = 0.74; figure 2a). Considering both and technical
differences between the tow methods and the biological changes along
the lifecycle this confirms the adequacy of our methods for assessing
the host transcriptome.

325 mouse genes were considered differently expressed (DE; FDR<1\%)
between uinfected and 3dpi, 1,804 between controls and 5dpi and 2,711
genes between 0dpi and 7dpi. This lead to combined set of 3453 genes
reacting to infection (figure 2bi).  DE genes early in infection (3dpi
and and 5dpi) were not simply a subset of genes becoming DE later in
infection (7dpi). To further validate this pattern of a distinct
response in early infection we performed hierarchical clustering on
(the union of) mouse genes DE between different dpi (figure 2c).

The changes in early infection are ...

Those changes are similarly pronounce in the immunocompetent strains
and RAG-/- immunodefficient animals.

The ponounced changes later in infection are is likely reflecting the
onset of an adaptive immune response.
